Categories of Optics

### 2018-03-02 paper.tex
Some attempts at diagrams

\usepackage{tikz-cd}
\usetikzlibrary{cd}
\usetikzlibrary{arrows}

@@

\title{The Categorical Structure of Optics}
\author{Mitchell Riley}
\begin{document}
\maketitle

@@

Let $\C$ be a symmetric monoidal category, and suppose we have two pairs of objects $(S, S')$ and $(A, A')$. An \emph{optic} $p : (S, S') \hto (A, A')$ is an element of the set
\begin{align*}
\Optic_\C((S, S'), (A, A')) := \int^{M \in \C} \C(S, M \otimes A) \times \C(M \otimes A', S')
\end{align*}
If $l : S \to M \otimes A$ and $r : M \otimes A' \to S'$, we write $\langle l, r \rangle : (S, S') \hto (A, A')$ for the image of this pair in $\Optic_\C((S, S'), (A, A'))$, and say that $M$ is the complement for this representative.

\begin{remark}
Optics will always be written with a crossed arrow $\hto$, to distinguish them from morphisms of $\C$.
\end{remark}

Elements of $\Optic_\C((S, S'), (A, A'))$ have a suggestive interpretation as string diagrams with a ``hole'' missing. We draw the pair $\langle l, r \rangle$ as
\begin{center}
\input{diagrams/generic-optic.tikz}
\end{center}
using the \todo{strange} convention of reading string diagrams from bottom to top.

@@

Graphically, this is placing the second optic into the hole of the first:
\begin{center}
\todo{todo}
\end{center}

@@

\begin{proposition}
For any $\C$, there is a functor $(-, -)_* : \C \times \C^\op \to \Optic_\C$, which is given on objects by $(S, S')_* = (S, S')$ and for a morphism $(f, g) : (S, S') \to (A, A')$ by $(f, g)_* = \langle \lambda_A^{-1} f, g \lambda_{A'} \rangle$.
\end{proposition}
\begin{proof}
This preserves identities, as the identity on an object $(S, S')$ in $\Optic_\M$ is defined to be exactly $\langle \lambda^{-1}_S, \lambda_{S'} \rangle$.

To check functoriality, suppose we have $(f, g) : (S, S') \to (A, A')$ and $(f', g') : (A, A') \to (B, B')$ in $\C \times \C^\op$. Then:
\begin{align*}
(f', g')_* \circ (f, g)_* 
&= \langle \lambda^{-1}_B f', g' \lambda_{B'} \rangle \circ \langle \lambda^{-1}_A f, g \lambda_{A'} \rangle \\
&= \langle (I\otimes (\lambda^{-1}_B f'))\lambda^{-1}_A f, g \lambda_{A'} (I\otimes (g' \lambda_{B'}))\rangle && \text{(By definition of $\circ$)}\\
&= \langle (I \otimes \lambda^{-1}_B) (I \otimes f')\lambda^{-1}_A f, g \lambda_{A'} (I \otimes g')(I\otimes \lambda_{B'})\rangle && \text{(Functoriality of $I \otimes -$)}\\
&= \langle (I\otimes \lambda^{-1}_B) \lambda^{-1}_B f' f, g g' \lambda_{B'} (I\otimes \lambda_{B'})\rangle && \text{(Naturality of $\lambda$)}\\
&= \langle (\lambda^{-1}_I \otimes B) \lambda^{-1}_B f' f, g g' \lambda_{B'} (\lambda_I \otimes B')\rangle && \text{(Unitality of action)} \\
&= \langle \lambda^{-1}_B f' f, g g' \lambda_{B'} (\lambda_I \otimes B') (\lambda^{-1}_I \otimes B') \rangle && \text{(Coend relation)}  \\
&= \langle \lambda^{-1}_B f'f, g g' \lambda_{B'} \rangle \\
&= (f'f, gg')_*
\end{align*}
Graphically, there is not much to do:
\begin{center}
\todo{todo}
\end{center}
\end{proof}

@@

$\Optic_\C$ is symmetric monoidal, where $(S, S') \otimes (T, T') = (S \otimes T, S' \otimes T')$, the unit is $(I, I)$, and with the action on morphisms induced by

@@

For future convenience, note that 
\langle l, r \rangle \otimes (T, T') &= \langle (M \otimes s_{A,I} \otimes T)(l \otimes \lambda_T), (r \otimes \lambda^{-1}_{T'})(M \otimes s_{A',I}\otimes T') \rangle \\
&= \langle (\lambda^{-1}_M \otimes A \otimes T)(M \otimes s_{A,I} \otimes T)(l \otimes \lambda_T), (r \otimes \lambda^{-1}_{T'})(M \otimes s_{A',I}\otimes T')(\lambda_M \otimes A' \otimes T') \rangle \\
&= \langle l \otimes T, r \otimes T' \rangle
Now we check functoriality. Suppose we have:
\langle l_1, r_1 \rangle : (S_1, S_1') &\hto (S_2, S_2') \\
\langle l_2, r_2 \rangle : (S_2, S_2') &\hto (S_3, S_3') \\
\langle p_1, q_1 \rangle : (T_1, T_1') &\hto (T_2, T_2') \\
\langle p_2, q_2 \rangle : (T_2, T_2') &\hto (T_3, T_3')
with complements $M_1, M_2, N_1, N_2$ respectively. Then: \todo {(This is a mess. String diagram?)}
&(\langle l_2, r_2 \rangle \otimes \langle p_2, q_2 \rangle) \circ (\langle l_1, r_1 \rangle \otimes \langle p_1, q_1 \rangle) \\
= \; &\langle (M_2 \otimes s_{S_3,N_2} \otimes T_3)(l_2 \otimes p_2), (r_2 \otimes q_2)(M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
&\circ \langle (M_1 \otimes s_{S_2,N_1} \otimes T_2)(l_1 \otimes p_1), (r_1 \otimes q_1)(M_1 \otimes s_{S_2',N_1} \otimes T_2') \rangle \\
= \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes N_1 \otimes l_2 \otimes p_2)(M_1 \otimes s_{S_2,N_1} \otimes T_2)(l_1 \otimes p_1), \\
&\;(r_1 \otimes q_1)(M_1 \otimes s_{S_2',N_1} \otimes T_2')(M_1 \otimes N_1 \otimes r_2 \otimes q_2)(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
= \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes s_{(M_2 \otimes S_3),N_1} \otimes T_3)(M_1 \otimes  l_2 \otimes N_1 \otimes p_2)(l_1 \otimes p_1), \\
&\;(r_1 \otimes q_1)(M_1 \otimes r_2 \otimes N_1 \otimes q_2)(M_1 \otimes s_{(M_2 \otimes S_3'),N_1} \otimes T_3')(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
= \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes s_{(M_2 \otimes S_3),N_1} \otimes T_3)((M_1 \otimes l_2)l_1\otimes (N_1 \otimes p_2)p_1), \\
&\;(r_1(M_1 \otimes r_2) \otimes q_1(N_1 \otimes q_2))(M_1 \otimes s_{(M_2 \otimes S_3'),N_1} \otimes T_3')(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
= \; & \text{\todo{(Add a $s_{M_2, N_1}$ to either side of the coend, then appeal to coherence)}} \\
= \; &\langle (M_1 \otimes M_2 \otimes s_{(N_1 \otimes N_2),S_3} \otimes T_3)((M_1 \otimes l_2)l_1\otimes (N_1 \otimes p_2)p_1) ,  \\
&\;(r_1(M_1 \otimes r_2) \otimes q_1(N_1 \otimes q_2))(M_1 \otimes M_2 \otimes s_{(N_1 \otimes N_2),S'_3} \otimes T'_3) \rangle \\
= \; &\langle (M_1 \otimes l_2)l_1, r_1(M_1 \otimes r_2) \rangle \otimes \langle (N_1 \otimes p_2)p_1, q_1(N_1 \otimes q_2) \rangle \\
= \; &(\langle l_2, r_2 \rangle  \circ \langle l_1, r_1 \rangle) \otimes (\langle p_2, q_2 \rangle \circ \langle p_1, q_1 \rangle)

The structure morphisms are all lifted from the structure morphisms in $\C$:
\begin{align*}
\alpha_{(R, R'), (S, S'), (T, T')} &:= (\alpha_{R,S,T}, \alpha_{R',S',T'}^{-1})_* \\
\lambda_{(S, S')} &:= (\lambda_{(S, S')}, \lambda_{(S, S')}^{-1})_* \\
\rho_{(S, S')} &:= (\rho_{(S, S')}, \rho_{(S, S')}^{-1})_* \\
s_{(S, S'), (T, T')} &:= (s_{S, T}, s_{T', S'})_*
\end{align*}

Note that because $(S, S')_* = (S, S')$, required equations for $(-, -)_*$ to be a monoidal functor hold by definition (although we don't yet know that $\Optic_\otimes$ is monoidal). The pentagon and triangle axioms then hold in $\Optic_\otimes$, as they are the image of the same axioms in $\C \times \C^\op$ under $(-, -)_*$. The only remaining piece to verify is that these structure maps are natural in $\Optic_\otimes$.

Let us check that $\alpha_{(R, R'), (S, S'), (T, T')}$ is natural in $(R, R')$. Suppose $f : (Q, Q') \hto (R, R')$ has representative $(l, r) : (Q \to M \otimes R, M \otimes R' \to Q')$, then:
\begin{align*}
&\alpha_{(R, R'), (S, S'), (T, T')} \circ (f \otimes (S, S')) \otimes (T, T') \\
&= (\alpha_{R,S,T}, \alpha_{R',S',T'}^{-1})_*  \circ (\langle l, r \rangle \otimes (S, S')) \otimes (T, T') \\
&= \langle\lambda_{R \otimes (S \otimes T)}\alpha_{R,S,T}, \alpha_{R',S',T'}^{-1} \lambda^{-1}_{(R' \otimes S') \otimes T'} \rangle  \circ \left\langle (l \otimes S) \otimes T, (r \otimes S') \otimes T' \right\rangle \\
&=\left\langle (M \otimes \lambda_{R \otimes (S \otimes T)}\alpha_{R,S,T})((l \otimes S) \otimes T), ((r \otimes S') \otimes T') (M \otimes \alpha_{R',S',T'}^{-1} \lambda^{-1}_{(R' \otimes S') \otimes T'} ) \right\rangle \\
&=\left\langle (M \otimes \alpha_{R,S,T})((l \otimes S) \otimes T), ((r \otimes S') \otimes T') (M \otimes \alpha_{R',S',T'}^{-1} )\right\rangle \\
&=\left\langle (l \otimes (S \otimes T))\alpha_{Q,S,T}, \alpha_{Q',S',T'}^{-1}(r \otimes (S' \otimes T')) \right\rangle \\
&= \langle (I \otimes l \otimes (S \otimes T)) \lambda_{Q \otimes (S \otimes T)}\alpha_{Q,S,T}, (I \times r \otimes (S' \otimes T')) \alpha_{Q',S',T'}^{-1} \lambda^{-1}_{Q' \otimes (S' \otimes T')}\rangle \\
&= \langle l \otimes (S \otimes T), r \otimes (S' \otimes T')\rangle \circ \left\langle \lambda_{Q \otimes (S \otimes T)}\alpha_{Q,S,T}, \alpha_{Q',S',T'}^{-1} \lambda^{-1}_{Q' \otimes (S' \otimes T')} \right\rangle \\
&= \langle l, r \rangle \otimes ((S, S') \otimes (T, T')) \circ (\alpha_{Q,S,T}, \alpha_{Q',S',T'}^{-1})_* \\
&= f \otimes ((S, S') \otimes (T, T')) \circ \alpha_{(Q, Q'), (S, S'), (T, T')} 
\end{align*}
\todo{There is some cheating going on here, composition actually adds an extra $\alpha$, but that wouldn't get in the way.}
\end{proof}

@@

There is some further useful structure in $\Optic_\C$. 
\begin{proposition}
The set of costates $(S, S') \hto (I, I)$ is isomorphic to $\C(S, S')$.
\end{proposition}
\begin{proof}
\begin{align*}
\Optic_\otimes((S, S'), (I, I)) 
&= \int^{M \in \C} \C(S, M \otimes I) \times \C(M \otimes I, S') \\
&\cong \int^{M \in \C} \C(S, M) \times \C(M, S') \\
&\cong \C(S, S')
\end{align*}
\end{proof}

States are not as easy to describe in general:

@@

\todo{Entrywise Coproduct}

\begin{proposition}
A strong monoidal functor $F : \C \to \D$ induces a functor $\Optic(F) : \Optic_\C \to \Optic_\D$
\end{proposition}
\begin{proof}
\todo{todo}
\end{proof}

\begin{proposition}
These assignments assemble into a \todo{pseudo?}functor $\Optic : \SymmMonCat \to SymmMonCat$.
\end{proposition}
\begin{proof}
\todo{todo}
\end{proof}

@@

As noted in \cite{CoherenceForLenses},

\begin{definition}[{\cite[Definition 5.1]{CoherenceForLenses}}]
A \emph{teleological category is} a symmetric monoidal category $(\C, \otimes, I)$, equipped with:
\begin{itemize}
\item A wide symmetric monoidal subcategory $\C_d$ of dualisable morphisms, with an involutive symmetric monoidal functor $(-)^* : \C_d^\op \to \C_d$
\item A family of morphisms $\epsilon_X : X \otimes X^* \to I$, called \emph{counits}, natural with respect to the morphisms in $\C_d$,
\end{itemize}
such that the diagrams
blah
commute.
\end{definition}

In the case of optics, 
The identity morphism $\C(S, S)$ gives us a preoptic $\varepsilon : (S, S) \to (I, I)$ that we call the \emph{counit}.

\begin{proposition}
With the above data, the category $\Optic_\C$ is a teleological category.
\end{proposition}
\begin{proof}
\todo{todo}
\end{proof}

We can therefore

As conjectured by Jules Hedges, the structure of a teleological category is precisely what is added by $\Optic$.
\begin{theorem}
$\Optic_\C$ is the free teleological category on a symmetric monoidal category $\C$.
\end{theorem}
\begin{proof}
\todo{todo}
\end{proof}

@@

Let $\C$ be a category and $(\M, \otimes, I)$ a monoidal category that acts on $\C$.

\section{Lawful Optics}
In this section we will focus only on preoptics of the form $p : (S,S) \hto (A, A)$. In what follows, we abbreviate $\Optic_\M((S, S), (A, A))$ as $\Optic_\M(S, A)$ and $p : (S, S) \hto (A, A)$ as $p : S \hto A$.

@@

\begin{proposition}
\label{prop-onthenose}
If $p : S \to A$ is an optic, then there exists a representative $p = \langle l, r \rangle$ such that $rl = \phi A$ on the nose for some $\phi : M \to M$.
\end{proposition}
\begin{proof}
\todo{This is painful but seems important. There may be a more general principle at work here but I couldn't see it.}

@@

@@

\subsection{Change of Action}

\todo{TODO}

>>>>

\usepackage{tikz-cd}
\usetikzlibrary{positioning}
\usetikzlibrary{fit}
\usetikzlibrary{cd}
\usetikzlibrary{arrows}

\usepackage{adjustbox}
\usepackage{etoolbox}
\BeforeBeginEnvironment {tikzpicture}{\begin{adjustbox}{valign=m}}
\AfterEndEnvironment {tikzpicture}{\end{adjustbox}}

@@

\title{Categories of Optics}
\author{Mitchell Riley}
\begin{document}
\maketitle

@@

Let $\C$ be a symmetric monoidal category, and suppose we have two pairs of objects $(S, S')$ and $(A, A')$. An \emph{optic} $p : (S, S') \hto (A, A')$ is an element of the set
\begin{align*}
\Optic_\C((S, S'), (A, A')) := \int^{M \in \C} \C(S, M \otimes A) \times \C(M \otimes A', S')
\end{align*}
If $l : S \to M \otimes A$ and $r : M \otimes A' \to S'$, we write $\langle l, r \rangle : (S, S') \hto (A, A')$ for the image of this pair in $\Optic_\C((S, S'), (A, A'))$, and say that $M$ is the complement for this representative. Optics will always be written with a crossed arrow $\hto$, to distinguish them from morphisms of $\C$.

\begin{remark}
Elements of $\Optic_\C((S, S'), (A, A'))$ have an appealing interpretation as string diagrams with a ``hole'' missing. We draw the pair $\langle l, r \rangle$ as
\begin{center}
\input{diagrams/generic-optic.tikz}
\end{center}
reading left to right, so the left half of the diagram represents $l$ and the right half $r$. The relation imposed by the coend takes the following form:
\begin{center}
\input{diagrams/coend-relation-left.tikz} \quad $=$ \quad \input{diagrams/coend-relation-right.tikz}
\end{center}
We will therefore omit the vertical cut point between $l$ and $r$ in subsequent diagrams; any choice yields a representative of same optic.
\end{remark}

@@

Graphically, this is placing the second optic into the middle of the first:
\begin{center}
\todo{todo}
\end{center}

@@

\begin{proposition}
For any $\C$, there is a functor $\iota : \C \times \C^\op \to \Optic_\C$, which is given on objects by $\iota(S, S') = (S, S')$ and for a morphism $(f, g) : (S, S') \to (A, A')$ by $\iota(f, g) = \langle \lambda_A^{-1} f, g \lambda_{A'} \rangle$.
\end{proposition}
\begin{proof}
This preserves identities, as the identity on an object $(S, S')$ in $\Optic_\M$ is defined to be exactly $\langle \lambda^{-1}_S, \lambda_{S'} \rangle$.

To check functoriality, suppose we have $(f, g) : (S, S') \to (A, A')$ and $(f', g') : (A, A') \to (B, B')$ in $\C \times \C^\op$. Then:
\begin{align*}
\iota(f', g') \circ \iota(f, g) 
&= \langle \lambda^{-1}_B f', g' \lambda_{B'} \rangle \circ \langle \lambda^{-1}_A f, g \lambda_{A'} \rangle \\
&= \langle (I\otimes (\lambda^{-1}_B f'))\lambda^{-1}_A f, g \lambda_{A'} (I\otimes (g' \lambda_{B'}))\rangle && \text{(By definition of $\circ$)}\\
&= \langle (I \otimes \lambda^{-1}_B) (I \otimes f')\lambda^{-1}_A f, g \lambda_{A'} (I \otimes g')(I\otimes \lambda_{B'})\rangle && \text{(Functoriality of $I \otimes -$)}\\
&= \langle (I\otimes \lambda^{-1}_B) \lambda^{-1}_B f' f, g g' \lambda_{B'} (I\otimes \lambda_{B'})\rangle && \text{(Naturality of $\lambda$)}\\
&= \langle (\lambda^{-1}_I \otimes B) \lambda^{-1}_B f' f, g g' \lambda_{B'} (\lambda_I \otimes B')\rangle && \text{(Unitality of action)} \\
&= \langle \lambda^{-1}_B f' f, g g' \lambda_{B'} (\lambda_I \otimes B') (\lambda^{-1}_I \otimes B') \rangle && \text{(Coend relation)}  \\
&= \langle \lambda^{-1}_B f'f, g g' \lambda_{B'} \rangle \\
&= \iota(f'f, gg')
\end{align*}
Graphically, there is not much to do:
\begin{center}
\todo{todo}
\end{center}
\end{proof}

@@

$\Optic_\C$ is symmetric monoidal, where $(S, S') \otimes (T, T') = (S \otimes T, S' \otimes T')$, the unit is $(I, I)$, and the action on morphisms induced by

@@

This is functorial, as suggested by the equivalence of the following string diagrams:
\todo{todo}
%For future convenience, note that 
%\begin{align*}
%\langle l, r \rangle \otimes (T, T') &= \langle (M \otimes s_{A,I} \otimes T)(l \otimes \lambda_T), (r \otimes \lambda^{-1}_{T'})(M \otimes s_{A',I}\otimes T') \rangle \\
%&= \langle (\lambda^{-1}_M \otimes A \otimes T)(M \otimes s_{A,I} \otimes T)(l \otimes \lambda_T), (r \otimes \lambda^{-1}_{T'})(M \otimes s_{A',I}\otimes T')(\lambda_M \otimes A' \otimes T') \rangle \\
%&= \langle l \otimes T, r \otimes T' \rangle
%\end{align*}
%Now we check functoriality. Suppose we have:
%\begin{align*}
%\langle l_1, r_1 \rangle : (S_1, S_1') &\hto (S_2, S_2') \\
%\langle l_2, r_2 \rangle : (S_2, S_2') &\hto (S_3, S_3') \\
%\langle p_1, q_1 \rangle : (T_1, T_1') &\hto (T_2, T_2') \\
%\langle p_2, q_2 \rangle : (T_2, T_2') &\hto (T_3, T_3')
%\end{align*}
%with complements $M_1, M_2, N_1, N_2$ respectively. Then: \todo {(This is a mess. String diagram?)}
%\begin{align*}
%&(\langle l_2, r_2 \rangle \otimes \langle p_2, q_2 \rangle) \circ (\langle l_1, r_1 \rangle \otimes \langle p_1, q_1 \rangle) \\
%= \; &\langle (M_2 \otimes s_{S_3,N_2} \otimes T_3)(l_2 \otimes p_2), (r_2 \otimes q_2)(M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
%&\circ \langle (M_1 \otimes s_{S_2,N_1} \otimes T_2)(l_1 \otimes p_1), (r_1 \otimes q_1)(M_1 \otimes s_{S_2',N_1} \otimes T_2') \rangle \\
%= \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes N_1 \otimes l_2 \otimes p_2)(M_1 \otimes s_{S_2,N_1} \otimes T_2)(l_1 \otimes p_1), \\
%&\;(r_1 \otimes q_1)(M_1 \otimes s_{S_2',N_1} \otimes T_2')(M_1 \otimes N_1 \otimes r_2 \otimes q_2)(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
%= \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes s_{(M_2 \otimes S_3),N_1} \otimes T_3)(M_1 \otimes  l_2 \otimes N_1 \otimes p_2)(l_1 \otimes p_1), \\
%&\;(r_1 \otimes q_1)(M_1 \otimes r_2 \otimes N_1 \otimes q_2)(M_1 \otimes s_{(M_2 \otimes S_3'),N_1} \otimes T_3')(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
%= \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes s_{(M_2 \otimes S_3),N_1} \otimes T_3)((M_1 \otimes l_2)l_1\otimes (N_1 \otimes p_2)p_1), \\
%&\;(r_1(M_1 \otimes r_2) \otimes q_1(N_1 \otimes q_2))(M_1 \otimes s_{(M_2 \otimes S_3'),N_1} \otimes T_3')(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
%= \; & \text{\todo{(Add a $s_{M_2, N_1}$ to either side of the coend, then appeal to coherence)}} \\
%= \; &\langle (M_1 \otimes M_2 \otimes s_{(N_1 \otimes N_2),S_3} \otimes T_3)((M_1 \otimes l_2)l_1\otimes (N_1 \otimes p_2)p_1) ,  \\
%&\;(r_1(M_1 \otimes r_2) \otimes q_1(N_1 \otimes q_2))(M_1 \otimes M_2 \otimes s_{(N_1 \otimes N_2),S'_3} \otimes T'_3) \rangle \\
%= \; &\langle (M_1 \otimes l_2)l_1, r_1(M_1 \otimes r_2) \rangle \otimes \langle (N_1 \otimes p_2)p_1, q_1(N_1 \otimes q_2) \rangle \\
%= \; &(\langle l_2, r_2 \rangle  \circ \langle l_1, r_1 \rangle) \otimes (\langle p_2, q_2 \rangle \circ \langle p_1, q_1 \rangle)
%\end{align*}

The structure morphisms are all lifted from the structure morphisms in $\C$:
\begin{align*}
\alpha_{(R, R'), (S, S'), (T, T')} &:= \iota(\alpha_{R,S,T}, \alpha_{R',S',T'}^{-1}) \\
\lambda_{(S, S')} &:= \iota(\lambda_{(S, S')}, \lambda_{(S, S')}^{-1}) \\
\rho_{(S, S')} &:= (\rho_{(S, S')}, \rho_{(S, S')}^{-1}) \\
s_{(S, S'), (T, T')} &:= \iota(s_{S, T}, s_{T', S'})
\end{align*}

Note that because $\iota(S, S') = (S, S')$, required equations for $\iota$ to be a monoidal functor hold by definition (although we don't yet know that $\Optic_\otimes$ is monoidal). The pentagon and triangle axioms then hold in $\Optic_\otimes$, as they are the image of the same axioms in $\C \times \C^\op$ under $\iota$. The only remaining thing to verify is that these structure maps are natural in $\Optic_\otimes$.

Let us check that $\alpha_{(R, R'), (S, S'), (T, T')}$ is natural in $(R, R')$. Suppose $f : (Q, Q') \hto (R, R')$ has representative $(l, r) : (Q \to M \otimes R, M \otimes R' \to Q')$, then:
\begin{align*}
&\alpha_{(R, R'), (S, S'), (T, T')} \circ (f \otimes (S, S')) \otimes (T, T') \\
&= \iota(\alpha_{R,S,T}, \alpha_{R',S',T'}^{-1})  \circ (\langle l, r \rangle \otimes (S, S')) \otimes (T, T') \\
&= \langle\lambda_{R \otimes (S \otimes T)}\alpha_{R,S,T}, \alpha_{R',S',T'}^{-1} \lambda^{-1}_{(R' \otimes S') \otimes T'} \rangle  \circ \left\langle (l \otimes S) \otimes T, (r \otimes S') \otimes T' \right\rangle \\
&=\left\langle (M \otimes \lambda_{R \otimes (S \otimes T)}\alpha_{R,S,T})((l \otimes S) \otimes T), ((r \otimes S') \otimes T') (M \otimes \alpha_{R',S',T'}^{-1} \lambda^{-1}_{(R' \otimes S') \otimes T'} ) \right\rangle \\
&=\left\langle (M \otimes \alpha_{R,S,T})((l \otimes S) \otimes T), ((r \otimes S') \otimes T') (M \otimes \alpha_{R',S',T'}^{-1} )\right\rangle \\
&=\left\langle (l \otimes (S \otimes T))\alpha_{Q,S,T}, \alpha_{Q',S',T'}^{-1}(r \otimes (S' \otimes T')) \right\rangle \\
&= \langle (I \otimes l \otimes (S \otimes T)) \lambda_{Q \otimes (S \otimes T)}\alpha_{Q,S,T}, (I \times r \otimes (S' \otimes T')) \alpha_{Q',S',T'}^{-1} \lambda^{-1}_{Q' \otimes (S' \otimes T')}\rangle \\
&= \langle l \otimes (S \otimes T), r \otimes (S' \otimes T')\rangle \circ \left\langle \lambda_{Q \otimes (S \otimes T)}\alpha_{Q,S,T}, \alpha_{Q',S',T'}^{-1} \lambda^{-1}_{Q' \otimes (S' \otimes T')} \right\rangle \\
&= \langle l, r \rangle \otimes ((S, S') \otimes (T, T')) \circ \iota(\alpha_{Q,S,T}, \alpha_{Q',S',T'}^{-1}) \\
&= f \otimes ((S, S') \otimes (T, T')) \circ \alpha_{(Q, Q'), (S, S'), (T, T')} 
\end{align*}
\todo{There is some cheating going on here, composition actually adds an extra $\alpha$, but that wouldn't get in the way.}
\end{proof}

@@

\begin{proposition}
Any isomorphism $p : (S, S') \hto (A, A')$ in $\Optic_\C$ is of the form $\iota(f, g)$ for two isomorphisms $S \to A$ and $A' \to S'$ in $\C$.
\end{proposition}
\begin{proof}

\end{proof}

There is some further useful structure in $\Optic_\C$. 
\begin{proposition}
\label{prop-costates}
The set of costates $(S, S') \hto (I, I)$ is isomorphic to $\C(S, S')$.
\end{proposition}
\begin{proof}
\begin{align*}
\Optic_\otimes((S, S'), (I, I)) 
&= \int^{M \in \C} \C(S, M \otimes I) \times \C(M \otimes I, S') \\
&\cong \int^{M \in \C} \C(S, M) \times \C(M, S') \\
&\cong \C(S, S')
\end{align*}
\end{proof}

States are not as easy to describe in general, however:

@@

\begin{proposition}
\todo{Entrywise Coproduct}
\end{proposition}

\begin{remark}
The motivating example here is that of coproducts distributing over the tensor product.
\end{remark}

\begin{proposition}
A strong monoidal functor $F : \C \to \D$ induces a functor $\Optic(F) : \Optic_\C \to \Optic_\D$
\end{proposition}
\begin{proof}
On objects this is simply $\Optic(F)(S, S') = (FS, FS')$. The action on morphisms $\Optic_\C((S, S'), (A, A'))$ is induced by
\begin{align*}
&\C(S, M \otimes A) \times \C(M \otimes A', S) \\
\to &\D(FS, F(M \otimes A)) \times \D(F(M \otimes A'), FS) \\
\to &\D(FS, FM \otimes FA) \times \D(FM \otimes FA', FS) \\
\to &\int^{N \in \D} \D(FS, N \otimes FA) \times \D(N \otimes FA', FS).
\end{align*}
where we have used the strong monoidal isomorphism in both directions.
\todo{use equations instead}
\todo{argue for functoriality}
\end{proof}

\begin{proposition}
Together this operation defines a \todo{pseudo?}functor $\Optic : \SymmMonCat \to \SymmMonCat$.
\end{proposition}
\begin{proof}
\todo{todo}
\end{proof}

@@

As noted in \cite{CoherenceForLenses}, \todo{the category of optics admits its own string diagram language}

\begin{definition}[{\cite[Definition 5.1]{CoherenceForLenses}}]
A \emph{teleological category is} a symmetric monoidal category $(\C, \otimes, I)$, equipped with:
\begin{itemize}
\item A wide symmetric monoidal subcategory $\C_d$ of dualisable morphisms, with an involutive strong symmetric monoidal functor $(-)^* : \C_d^\op \to \C_d$
\item A symmetric monoidal extranatural family of morphisms $\varepsilon_X : X \otimes X^* \to I$, called \emph{counits}, natural with respect to the morphisms in $\C_d$.
\end{itemize}
\end{definition}
\todo{The involution bit seems to be almost the same as asking for a symmetric monoidal dagger-category, that is stricter than this though}

Unpacking the definition, $\varepsilon$ being a monoidal extranatural transformation amounts to the following diagrams in $\C$ commuting:
\[
\begin{tikzcd}
X \otimes Y^* \ar[r, "f \otimes Y^*"]  \ar[d, "X \otimes f^*", swap] & Y \otimes Y^* \ar[d, "\varepsilon_Y"] \\
X \otimes X^* \ar[r, "\varepsilon_X", swap] & I
\end{tikzcd} \hspace{1cm}
\begin{tikzcd}
X^* \otimes X \ar[r, "\sigma"]  \ar[dr, "\varepsilon_{X^*}", swap] & X \otimes X^* \ar[d, "\varepsilon_X"] \\
& I
\end{tikzcd} \hspace{1cm}
\begin{tikzcd}
X \otimes Y \otimes X^* \otimes Y^* \ar[r, "X \otimes s \otimes X^*"]  \ar[dr, "\varepsilon_{X \otimes Y}", swap] & X \otimes X^* \otimes Y \otimes Y^* \ar[d, "\varepsilon_X \otimes \varepsilon_Y"] \\
& I
\end{tikzcd}
\] 
where $f : X \to Y$ is in $\C_d$. \todo{Is that last one actually something extra? It seems like the duality should also reverse the order of tensor, or something}

\begin{proposition}
$\Optic_\C$ forms a teleological category, with the following data:
\begin{itemize}
\item The dualisable morphisms are all morphisms of the form $\iota(f, g)$;
\item The involution is given on objects by $(S, S')^* = (S', S)$, and on a morphism $\iota(f, g)$ by $\iota(g, f)$;
\item The counit $\varepsilon_{(S, S')}$  is given by the image of the symmetry $s_{S \otimes S'}$ under the isomorphism of Proposition \ref{prop-costates}.
\end{itemize}
\end{proposition}
\begin{proof}
That morphisms of the form $\iota(f, g)$ constitute a category and that $\iota$ is a symmetric monoidal involution is clear.

To check extranaturality, suppose we have a dualisable optic $\iota(f, g) : (X, X') \to (Y, Y')$, so $f : X \to Y$ and $g : Y' \to X'$. Extranaturality is witnessed by the equality of the string diagrams:
\begin{center}
\includegraphics[width=.4\textwidth]{diagrams/counit-extranaturality}
\end{center}
Symmetry:
\begin{center}
\includegraphics[width=.4\textwidth]{diagrams/counit-symmetry}
\end{center}
Monoidal:
\begin{center}
\includegraphics[width=.4\textwidth]{diagrams/counit-monoidal}
\end{center}
\end{proof}

\todo{They stress about the empty set, does this matter here?}

As conjectured by Jules Hedges, the structure of a teleological category is precisely what is added by forming the $\Optic$ category.
\begin{theorem}
$\Optic_\C$ is the free teleological category on a symmetric monoidal category $\C$.
\end{theorem}
\begin{proof}
\todo{todo}
\end{proof}

@@

To capture the \texttt{Iso}s, \texttt{Traversal}s and \texttt{Setter}s of the Haskell \texttt{lens} library, we generalise to the case of a monoidal action of one category on another.

\begin{definition}
Let $\C$ be a category and $(\M, \otimes, I)$ a monoidal category. An \emph{action of $\M$ on $\C$} is a strong monoidal functor $a : \M \to [\C, \C]$. The action of an object $M \in \M$ on $A \in \C$ is written $MA$.
\end{definition}

Given such an action, we define 
\begin{align*}
\Optic_\M((S, S'), (A, A')) := \int^{M \in \C} \C(S, MA) \times \C(MA', S')
\end{align*}

\begin{proposition}
We have a category $\Optic_\M$ and a functor $\iota : \C \times \C^\op \to \Optic_\M$. \qed
\end{proposition}

\subsection{Polymorphic Optics}

\section{Lawful Optics}
\todo{intro}

\begin{remark}
In this section we will focus only on preoptics of the form $p : (S,S) \hto (A, A)$. We therefore abbreviate $\Optic_\M((S, S), (A, A))$ as $\Optic_\M(S, A)$ and $p : (S, S) \hto (A, A)$ as $p : S \hto A$.
\end{remark}

@@

\begin{proposition}
There is an subcategory $\Lawful_\C$ of $\Optic_\C$ given by objects of the form $(X, X)$ and lawful optics between them.
\end{proposition}
\begin{proof}
The identity preoptic $\iota(\id_X, \id_X)$ is clearly an optic.

\begin{proposition}
\label{prop-onthenose}
If $p : S \to A$ is a lawful optic, then there exists a representative $p = \langle l, r \rangle$ such that $rl = \phi A$ on the nose for some $\phi : M \to M$.
\end{proposition}
\begin{proof}
\todo{This is painful but seems important. There may be a more general principle at work here but I couldn't see it.}

@@

\subsubsection{Isos}

@@
